Fast Sorting on Modern Computers

### 2005-06-16 docs/tech_report/merge.tex
fixed a demo script that was overwriting a merge result fixed the writing in merge sort for the level 1 results

Knuth's algorithm S is a \n{straight} merge, and works as described above. It has
15\% lower instruction count than Algorithm N\footnote{This is observed on
random data, which algorithm N is not designed for. On more regular data, or
semi-sorted data, this will probably not occur.}. Algorithm S, like algorithm N,
is a very difficult algorithm to modify. Knuth describes it in assembly, and
making the optimizations that LaMarca describes on it is more difficult than
rewriting the algorithm more simply. The rewritten algorithm, referred to from
here on as `base mergesort', is a simple higher level version of algorithm S,
and performs only slightly more slowly.

@@

\section{Double Tiled Mergesort}
LaMarca's first problem with mergesort is the placement of the arrays with
respect to each other. If the source and target arrays map to the same cache
block, the writing from one to the other will cause a conflict miss. The next
read will cause another conflict miss, and will continue in this manner. To
solve this problem, the two arrays are positioned so that they map to different
parts of the cache. To do this, an additional amount of memory the same size as
the cache is allocated, which can be expensive for small arrays. The address of
the auxiliary array is then offset such that $index(source) + index(auxiliary) =
C$.

%TODO what does this formula mean. look up LaMarcas paper

@@

The final improvement was to fully sort keys in the level 1 cache, before
continuing to level 2 cache as in the tiled mergesort. The expected improvement
is not as significant as in the case of the level 2 cache, since the penalty
for level 1 misses is not great. For this reason, this sort is called Double
Tiled Mergesort instead of Tiled Mergesort, however, it is referred to as both
in this report.

Double Multi-mergesort has two phases. The first is the same as tiled mergesort, and
sorts the array into lists sized $C/2$. The second phase attempts to improve
temporal locality by merging these lists together in a single merge step. There
are $k$ lists and this is called a \n{k-way merge}. It is possible to do a
search across each of these keys, maintaining a list of indices to the smallest
unused key from each list. When there is a small number of lists, this isn't a
problem, however, when the number of lists increases significanty, the overhead
involved in this is substantial. It is necessary to check each of the lists, and
each key searched would cause a cache miss, since each array segment is aligned,
and map to the same cache block.

@@

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8k, so 2048 keys can be sorted in the level 1 cache.
262144 keys can be sorted in the level 2 cache. When arrays were pre-sorted,
they were pre-sorted to a size four, or eight if the number of passes was going
to be odd. These were reversed for the double multi-mergesort; since it does a
k-way merge into the source array, the sorted cache sized segments need to be in
the auxiliary array. An 8-heap was used for the multi-mergesorts.

@@

Algorithm N and algorithm S are included in the results for reference, but are
not discussed here.

The instruction count of mergesort is $O(NlogN)$, and does not vary
according to input. Multi-mergesort should have a significantly higher
instruction count once its threshold is hit.

The number of memory references is also $O(NlogN)$, and mergesort's cache
performance should be very regular, so long as the data set fits in the cache. Once
the data set exceeds the cache, the number of cache misses should leap. For
multi-mergesort, they should flatten out after this, perhaps with a slight
slope to take into account the conflict misses of the aligned array segments. The
fully associative caches should not have these misses, and the results should be
correspondingly lower. Due to their design, the double mergesorts should have
better performance than their counterparts.

@@

\plot{merge}{}{}{Except where marked, results shown are for a Direct Mapped cache.}{
Results shown are for a Direct Mapped cache.}{}{}

The number of instructions, shown in Figure \ref{merge instructions} is exactly
as predicted, with a 50\% spike at multi-mergesort. Level 2 misses, shown in
Figure \ref{merge level 2 misses} are also as expected, although the
multi-mergesort does not show the same improvement over tiled mergesort as in
LaMarca's results.

%TODO look over lamarcas results
The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
The reason for this is not clear.

Despite the improvement in cache misses due to multi-mergesort, it still cannot
account for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

The results presented here differ from LaMarca's results. In the case of
instruction count, both sets are nearly identical: the graphs are the same shape
and have similar ratios between the base and multi-mergesorts. The cache results
have a similar shape to LaMarca's, with a large increase in misses as soon as
the arrays no longer fit in the cache, and similar ratios between the base and
tiled versions.  But while LaMarca reports a very small increase in level 2
misses once multi-mergesort's data set is larger than the cache, the results
here show nearly the same jump as tiled mergesort. This results in half as many
misses as the base mergesort, while LaMarca shows nearly 10 times fewer misses
for his multi-mergesort.

As a result of this discrepancy, it is not a surprise that LaMarca's results
indicate that multi-mergesort is far faster than base-mergesort, but the same
results are not shown here. The increase in instruction count cripples
multi-mergesort, and tiled mergesort emerges the victor.

This is not to say that LaMarca's results disagree with those here. In fact, it
is noticeable that multi-mergesort's cache misses plateau at the edge of the
graph, while other mergesorts' misses continue to climb. This is in keeping with
both LaMarca's theory and results.

>>>>

Knuth's algorithm S is a \n{straight} merge, and works as described above. It has
15\% lower instruction count than Algorithm N\footnote{This is observed on
random data, which algorithm N is not designed for. On more regular data, or
semi-sorted data, this will probably not occur.}. Algorithm S, like algorithm N,
is a very difficult algorithm to modify. Knuth describes it in assembly, and
making the optimizations that LaMarca describes on it is more difficult than
rewriting the algorithm more simply. The rewritten algorithm, referred to from
here on as `base mergesort', is a simple higher level version of algorithm S,
and should perform only slightly more slowly.

@@

\section{Tiled Mergesort}
LaMarca's first problem with mergesort is the placement of the arrays with
respect to each other. If the source and target arrays map to the same cache
block, the writing from one to the other will cause a conflict miss. The next
read will cause another conflict miss, and will continue in this manner. To
solve this problem, the two arrays are positioned so that they map to different
parts of the cache. To do this, an additional amount of memory the same size as
the cache is allocated, which can be expensive for small arrays. The address of
the auxiliary array is then offset such that the block index in the source array
is $CacheSize/2$ from the block index in the target array. The is achieved by
inverting the most significant bit of the block index in the target array.

@@

\section{Double Tiled Mergesort}
Offsetting the target array to avoid conflicts reduces significantly the number
of level 2 cache misses. However, it does this at the expense of increasing the
number of level 1 cache misses. To avoid this, we furthur offset the target
array by $Level1CacheSize/2$. It is expected that the number of level 1 misses
should be greatly decreased, at the expense of very few level 2 misses. The
expected improvement is not as great as between base mergesort and tiled
mergesort, but some improvement should be seen.

\section{Multi-mergesort}

Multi-mergesort has two phases. The first is the same as tiled mergesort, and
sorts the array into lists half the size of the cache. The second phase attempts
to improve temporal locality by merging these lists together in a single merge
step. There are $k$ lists and this is called a \n{k-way merge}. It is possible
to do a search across each of these keys, maintaining a list of indices to the
smallest unused key from each list. When there is a small number of lists, this
isn't a problem, however, when the number of lists increases significanty, the
overhead involved in this is substantial. It is necessary to check each of the
lists, and each key searched would cause a cache miss, since each array segment
is aligned, and map to the same cache block.

@@

Double Multi-mergesort combines the double tiling of Double Tiled mergesort with
Multi-mergesort.

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
pre-sorted, they were pre-sorted to a size four, or eight if the number of
passes was going to be odd. These were reversed for the double multi-mergesort;
since it does a k-way merge into the source array, the sorted cache sized
segments need to be in the auxiliary array. An 8-heap was used for the
multi-mergesorts.

@@

Algorithm N is not optimized for random data, and should not perform as well as
algorithm S. Algorithm S is expected to have a lower instruction count than base
mergesort. Neither algorithm is expected to have good cache performance,

The instruction count of mergesort is $O(NlogN)$, and does not vary
according to input. Multi-mergesort should have a significantly higher
instruction count once its threshold is hit. The double merges have exactly the
same flow control as the single merges, so will only differ in cache misses.

The number of memory references is also $O(NlogN)$, and mergesort's cache
performance should be very regular, so long as the data set fits in the cache. Once
the data set exceeds the cache, the number of cache misses should leap. For
multi-mergesort, they should flatten out after this, perhaps with a slight
slope to take into account the conflict misses of the aligned array segments. The
fully associative caches should not have these misses, and the results should be
correspondingly lower. Due to their design, the double mergesorts should have
better performance than their counterparts, both in level 1 misses, and overall,
though the difference is not expected to be great.

@@

\plot{merge}{}{}{Results shown are for a Direct Mapped cache.}{
Results shown are for a Direct Mapped cache.}{}{}

As expected, algorithm N has lower instruction count than algorithm S, which
also slightly outperforms base mergesort.  The number of instructions, shown in
Figure \ref{merge instructions} is exactly as predicted, with a 50\% spike at
multi-mergesort. Level 2 misses, shown in Figure \ref{merge level 2 misses} are
also as expected, although the multi-mergesort does not show the same
improvement over tiled mergesort as in LaMarca's results.

The prediction results in Figure \ref{Branch Simulation results for mergesort}
are as expected. Very little variation exists between the types of mergesort. It
is noticeable, though, that the spike in branch misses due to the
multi-mergesorts is considerably lessened by the two level adaptive predictor.
The reason for this is not clear.

As predicted, the rate of level 1 misses is greatly reduced by the double
tiling, as shown in Figure \ref{merge level 1 misses}. Double multi-mergesort
has consistantly four times fewer level 1 misses than multi-mergesort, and both
double tiled sorts perform significantly better than the unoptimizied sorts.

The level 2 cache results are startingly similar to LaMarca's. The shape of the
Figure \ref{merge level 2 misses} is exactly that provided by LaMarca. The
ratio's between the each of the sorts is also the same, though the numbers are
not identical. As we use a different base algorithm from LaMarca, this is not
unthinkable.

Despite the improvement in cache misses due to multi-mergesort, it still cannot
compensate for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

The cache misses end up being the deciding factor between the performance of
double tiled mergesort and base mergesort, the next best contender. Dispite the
reduction in level 2 cache misses, the sharp increase in level 1 misses result
is poor performance for tiled and multi-mergesort. Multi-mergsort is further
encumbered with a much higher instruction count. Double tiled mergesort, with
its low instruction count, and low cache miss rate in both level 1 and two,
emerges the winner.
